Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-31 cis-perfect-testing.tex
proof

In survival analysis, the interest is in estimating the distribution of the interval between an \emph{initiating event} and a \emph{terminating event}.
In the current context, the initiating event is the episode's start; the terminating event is the episode's end.
I start by introducing the relevant, standard concepts in survival analysis; a wide range of texts cover them including \textcite[chapter 1]{bogaertsSurvival}.

@@

There are also undetected infections in the CIS (see \cref{perf-test:sec:undetected}).
However, the situation is subtly different from the standard truncation situation encountered in the literature.
In the standard situation, nothing is known about the unrecorded intervals.
But in CIS, individuals in which undetected infections occurred are enrolled in the study.
Therefore, we know the times that they were tested.
This provides information on how long their infections may have occurred.
I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.

@@

Unlike in \textcite{heiseyModelling}, I must consider each individual's probability of undetected episodes separately because they are tested at different times.
In addition, I include the information from the individuals without a detected episode.
Furthermore, I marginalise the posterior density over the number of undetected episodes.
This makes inference tractable using the large sample size and the computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
% In particular, the posterior can be expressed within modern, high-performance Bayesian software such as Stan.

@@

An individual $i$ has a \emph{test schedule} $\sched_i$.
The test schedule $\sched_i$ is the set of times individual $i$ is tested at, starting from their last test prior to time 1, if it exists, or their time of enrolment otherwise.
Note that $\sched_i$ includes any tests that occur after $T$.
For example, for the individual in \cref{perf-test:fig:double-interval-censor}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
Each individual has exactly one test schedule, but it is possible that $\sched_i = \sched_{i'}$ for $i \neq i'$.
I assume that the test schedules are uninformative on all quantities of interest because they are part of the study design.
Therefore, I condition on them implicitly in all probabilities that follow.

@@

Now define the probabilities of each possible outcome of $O(j)$.
For detected outcomes, $\vec{\nu}_k \in \set{E}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$.
For undetected outcomes, let $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
Together, $\vec{p} = [p_1, \dots, p_D, p_u]^T$.
Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.
Then:
\begin{align}
  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
\intertext{that is:}
  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_D} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_D} p_k^{n_k}.
  \label{perf-test:eq:multinomial-ll}
\end{align}
\begin{figure}
\makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{cis-perfect-testing/regions_diag}}
\thisfloatpagestyle{empty}
\caption[Episode regions]{%
  Each dot is a combination of $b_j$ and $e_j$ for an arbitrary individual $i$.
  Each box, bounded by dashed lines, are combinations giving $O(j) = \nu_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
  $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
  The purple region corresponds to the doubly interval censored episode in this individual.
  That is, $n_8 = 1$ and $n_k = 0$ for $k = 1, \dots, 7, 9, \dots, 15$.
  The red region corresponds to combinations giving $O(j) = \emptyset$.
  Impossible region violates $b_j \leq e_j$.
}
\label{perf-test:fig:partitionSpace}
\end{figure}

@@

Decompose $p_k$ as $p_k = p_{ik} \prob(i(j) = i_k \mid \vec{\theta})$
where $p_{ik} = \prob(O(j) = \nu_k \mid i(j) = i_k, \vec{\theta})$.
% This is valid as $\prob(O(j) = \nu_k \mid i(j) \neq i_k) = 0$ due to the condition here being equivalent to equating the two vectors' final elements.
Assume that each infection episode occurs independently and with equal probability in any individual, \ie $\prob(i(j) = i_k) = 1/\Ncis$ for all $j$ and $k$.

@@

% An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
% Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
\begin{align}
\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t
\label{perf-test:eq:tau-def}
\end{align}
% defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
The first condition can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
% The fourth condition can be expressed as $\tau_{\sched_{i(j)}}(b_j) < \infty$.

@@

In this section I specify the form of $S_{\vec{\theta}}(t)$.
I parameterize $S$ in terms of the hazard $S_{\vec{\theta}}(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$ (see \cref{E-inc-prev:sec:definitions}).
Therefore, $\vec{\theta} = [\lambda_1, \dots, \lambda_{\dmax-1}]^T$.

@@

RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some normally distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ specify the prior on the initial gradient, $\lambda_2 - \lambda_1$; and $\mu_\sigma$ specifies the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/rw2-prior}
  \caption[RW2 prior for the hazard]{The prior predictive survival time (mean and 95\% central interval) for the RW2 prior specified in the main text. The posterior mean estimate using ATACCC data in \cref{E-ATACCC} is shown for comparison. \label{perf-test:fig:rw2-prior}}
\end{figure}

@@

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h}$ is the hazard, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{ATACCC:sec:duration}.
Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.
% The logit transformation maps from the $[0, 1]$ interval to the full real line.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  % \vspace{-1cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15. The density for smaller hazards is concentrated very close to 0. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
  \caption[Approximating the ATACCC posterior survival as a logit-normal]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

Having approximated the estimate as a multivariate normal, I add additional uncertainty using a discrete Beta process.
The discrete Beta process prior~\autocite{ibrahimBayesian,sunStatisticala} generalises the form of prior used in \cref{perf-test:sec:independent-priors} by allowing the central estimate of the hazard to vary over time.
It is:
\begin{align}
  \logit \vec{h} &\dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A) \\
  \lambda_t &\dist \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
  \alpha_t &= k_t h_t + \alpha_0 \\
  \beta_t &= k_t (1 - h_t) + \beta_0
\end{align}
where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and, as before, $h_t$ is an estimate of the hazard at time $t$ from \cref{E-ATACCC}.
An intuition for what this distribution represents derives from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).

@@

I simulate a dataset of detected episodes with the same characteristics as that in the CIS by the following procedure.
\begin{enumerate}
    \item Extract the test schedules for each individual who had at least one test during the period of interest.
    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2 July 2020 (100 days before the period where a detected episode would be included) and 6 December 2020 (the end of this period).
    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j} = b_{j} + d_i - 1$.
    \item Simulate the test results based on the test schedule, $b_{j}$, and $e_{j}$. Tests between $b_{j}$ and $e_{j}$ (inclusive) are positive, all other tests are negative.
    \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
    \item Of these remaining episodes, sample \numprint{4800} to match the sample size of the true dataset. This is needed because in step 2 the entire cohort was infected, while in the real study only an (unknown) portion is infected.
    \item For this final set of episodes, calculate $(l_j^{(b)}, r_j^{b}, l_j^{(e)}, r_j^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
\end{enumerate}

Step 3 requires a distribution of the duration of detectability.
I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to be consistent with the CIS.
I modify the tail based on an unpublished analysis of CIS data~\citePersonalComms{A.\ Sarah Walker}.
Walker uses survival analysis to estimate the duration from CIS data.
This analysis assumes the initiating event is known, and equal to the episode’s detection time. 
It assumes the final event is interval censored between the time of the final positive test and the subsequent negative test, or right censored if a negative test has not yet been observed.
A flexible, spline-based form is used for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.

@@

@@

I implemented the inference in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
Convergence was assessed using Rhat and ESS as described in \cref{E-MCMC:sec:convergence}, and all runs checked for divergent transitions.

The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.
This meant that the prior was very diffuse.

@@

Reasonable choices of the prior for the hazard are vague (although not flat) or informative.
A smoothing prior performed poorly, although that may be influenced by the choice of hyperparameters.
Reducing the amount of smoothing or type of smoothing may help.
More flexible smoothing priors could be used, such as a Gaussian Process for additional flexibility and smoothing and different timescales~\autocite{saulGaussian}.

The derivation in \cref{perf-test:sec:model} relied on an assumption that $O(j)$ and $O(j')$ occur independently for $j \neq j'$.
This is not the case, as immunity means that an individual previously infected is less likely to be infected in the future.
Additionally, the same individual cannot have multiple concurrent infections, due to the way that the episodes dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}) limiting certain combinations from occurring.
In practice, this means that $n_k > 1$ cannot occur.
The simulation study violated this assumption to a large extent: each individual could only have one infection.
Despite this, the method performed well, recovering the true survival function.
Therefore, it is unlikely that this assumption matters, possibly due to the large number of individuals without detected episodes.
Further work could explore different assumptions, \eg a ``full immunity'' assumption meaning each individual has zero or one episode in the period.

Various extensions could be worth exploring.
Most importantly, including covariates (\eg age, immunity from vaccination and/or previous infection, or variant).
Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocite{hakkiOnset,russellWithinhost}.
Covariates can be included with a variety of models, the simplest being a proportional hazards model.
A major challenge here is that the calculation of $p_t$ is greatly complicated because it is no longer linear in the survival function.

>>>>

In survival analysis, the interest is in estimating the distribution of the interval between an \emph{initiating event} and a \emph{terminating event}.
In the current context, the initiating event is the episode's start; the terminating event is the episode's end.
I start by introducing the relevant, standard concepts in survival analysis; a wide range of texts cover these concepts including \textcite[chapter 1]{bogaertsSurvival}.

@@

There are also undetected infections in the CIS (see \cref{perf-test:sec:undetected}).
However, the situation is subtly different from the standard truncation situation encountered in the literature.
In the standard situation, nothing is known about the unrecorded intervals.
But in CIS, individuals in which undetected infections occurred are enrolled in the study.
Therefore, we know the times that they were tested.
This provides information on how long their infections may have been.
I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.

@@

Unlike in \textcite{heiseyModelling}, I must consider each individual's probability of undetected episodes separately because they are tested at different times.
In addition, I include the information from the individuals without a detected episode.
Furthermore, I marginalise the posterior density over the number of undetected episodes.
This makes inference tractable using the large sample size and limited computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
% In particular, the posterior can be expressed within modern, high-performance Bayesian software such as Stan.

@@

An individual $i$ has a \emph{test schedule} $\sched_i$.
The test schedule $\sched_i$ is the set of times individual $i$ is tested at, starting from their last test prior to time 1, if it exists, or their time of enrolment otherwise; note that $\sched_i$ includes any tests that occur after $T$.
For example, for the individual in \cref{perf-test:fig:double-interval-censor}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
Each individual has exactly one test schedule, but it is possible that $\sched_i = \sched_{i'}$ for $i \neq i'$.
I assume that the test schedules are uninformative on all quantities of interest because they are part of the study design.
Therefore, I condition on them implicitly in all probabilities that follow.

@@

Now define the probabilities of each possible outcome of $O(j)$.
For detected outcomes, \ie $\vec{\nu}_k \in \set{E}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$.
For undetected outcomes, let $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
Together, $\vec{p} = [p_1, \dots, p_D, p_u]^T$.
Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.

Then:
\begin{align}
  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
\intertext{that is:}
  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_D} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_D} p_k^{n_k}.
  \label{perf-test:eq:multinomial-ll}
\end{align}
\begin{figure}
\makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{cis-perfect-testing/regions_diag}}
\thisfloatpagestyle{empty}
\caption[Episode regions]{%
  Each dot is a combination of $b_j$ and $e_j$ for an arbitrary individual $i$.
  Each box, bounded by dashed lines, are combinations giving $O(j) = \nu_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
  $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
  The purple region corresponds to the doubly interval censored episode in this individual.
  That is, $n_8 = 1$ and $n_k = 0$ for $k = 1, \dots, 7, 9, \dots, 15$.
  The red region corresponds to combinations giving $O(j) = \emptyset$.
  Impossible region violates $b_j \leq e_j$.
}
\label{perf-test:fig:partitionSpace}
\end{figure}

@@

Decompose $p_k$ as $p_k = p_{ik} \prob(i(j) = i_k \mid \vec{\theta})$
where $p_{ik} = \prob(O(j) = \vec{\nu}_k \mid i(j) = i_k, \vec{\theta})$.
% This is valid as $\prob(O(j) = \nu_k \mid i(j) \neq i_k) = 0$ due to the condition here being equivalent to equating the two vectors' final elements.
Assume that each infection episode occurs independently and with equal probability in any individual, \ie $\prob(i(j) = i_k) = 1/\Ncis$ for all $j$ and $k$.

@@

% An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
% Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
First define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
\begin{align}
\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t
\label{perf-test:eq:tau-def}
\end{align}
% defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
The first condition can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
% The fourth condition can be expressed as $\tau_{\sched_{i(j)}}(b_j) < \infty$.

@@

In this section I specify the form of $S_{\vec{\theta}}(t)$.
I parameterize $S$ in terms of the hazard $S_{\vec{\theta}}(t) = \prod_{i=1}^{t-1} (1 - \lambda_{i})$ (see \cref{E-inc-prev:sec:definitions}).
Therefore, $\vec{\theta} = [\lambda_1, \dots, \lambda_{\dmax-1}]^T$.

@@

An RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some normally distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ specify the prior on the initial gradient, $\lambda_2 - \lambda_1$; and $\mu_\sigma$ specifies the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/rw2-prior}
  \caption[RW2 prior for the hazard]{The prior predictive survival time (mean and 95\% central interval) for the RW2 prior specified in the main text. The posterior mean estimate using ATACCC data in \cref{E-ATACCC} is shown for comparison. \label{perf-test:fig:rw2-prior}}
\end{figure}

@@

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h}$ is the hazard, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{ATACCC:sec:duration}.
Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.
% The logit transformation maps from the $[0, 1]$ interval to the full real line.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  % \vspace{-1cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15. The density for smaller hazards is concentrated very close to 0. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
  \caption[Approximating the ATACCC posterior survival]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

Having approximated the estimate as a multivariate normal, I add additional uncertainty using a discrete Beta process.
The discrete Beta process prior~\autocite{ibrahimBayesian,sunStatisticala} generalises the form of prior used in \cref{perf-test:sec:independent-priors} by allowing the central estimate of the hazard to vary over time.
It is:
\begin{align}
  \logit \vec{h} &\dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A) \\
  \lambda_t &\dist \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
  \alpha_t &= k_t h_t + \alpha_0 \\
  \beta_t &= k_t (1 - h_t) + \beta_0
\end{align}
where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters.
An intuition for what this distribution represents derives from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).

@@

I simulate a dataset of detected episodes with the same characteristics as that in the CIS by the following procedure.
\begin{enumerate}
    \item Extract the test schedules for each individual who had at least one test during the period of interest.
    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2 July 2020 ($t = -100$) and 6 December 2020 ($t = T$).
    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j} = b_{j} + d_i - 1$.
    \item Simulate the test results based on the test schedule, $b_{j}$, and $e_{j}$. Tests between $b_{j}$ and $e_{j}$ (inclusive) are positive, all other tests are negative.
    \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
    \item Of these remaining episodes, sample \numprint{4800} to match the sample size of the true dataset. This is needed because in step 2 the entire cohort was infected, while in the real study only an (unknown) portion is infected.
    \item For this final set of episodes, calculate $(l_j^{(b)}, r_j^{b}, l_j^{(e)}, r_j^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
\end{enumerate}

Step 3 requires a distribution of the duration of detectability.
I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to be consistent with the CIS.
I modify the tail based on an unpublished analysis of the CIS data~\citePersonalComms{A.\ Sarah Walker}.
Walker uses survival analysis to estimate the duration from CIS data.
This analysis assumes the initiating event is known, and equal to the episode’s detection time. 
It assumes the final event is interval censored between the time of the final positive test and the subsequent negative test, or right censored if a negative test has not yet been observed.
A flexible, spline-based form is used for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.

@@

The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.

@@

I implemented the inference in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
Convergence was assessed using Rhat and ESS as described in \cref{E-MCMC:sec:convergence}, and all runs checked for divergent transitions.
This meant that the prior was very diffuse.

@@

Reasonable choices of the prior for the hazard are vague (although not flat) or informative.
A smoothing prior performed poorly, although that may be influenced by the choice of hyperparameters.
Reducing the amount of smoothing or changing the type of smoothing may help.
More flexible smoothing priors could be used, such as a Gaussian Process for additional flexibility and smoothing and different timescales~\autocite{saulGaussian}.

The derivation in \cref{perf-test:sec:model} relied on an assumption that $O(j)$ and $O(j')$ occur independently for $j \neq j'$.
This is not the case, as immunity means that an individual previously infected is less likely to be infected in the future.
Additionally, the same individual cannot have multiple detected concurrent infections, due to the way that the episodes dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}); this limits certain combinations from occurring.
In practice, this means that $n_k > 1$ cannot occur.
The simulation study violated this assumption to a large extent: each individual could only have one infection.
Despite this, the method performed well, recovering the true survival function.
Therefore, it is unlikely that this assumption matters, possibly due to the large number of individuals without detected episodes.
Further work could explore different assumptions, \eg a ``full immunity'' assumption meaning each individual has zero or one episode in the period.

Various extensions could be worth exploring.
Most importantly, including covariates (\eg age, immunity from vaccination and/or previous infection, or variant).
Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocite{hakkiOnset,russellWithinhost}.
Covariates can be included with a variety of models, the simplest being a proportional hazards model.

### 2024-03-31 conclusion.tex
proof

In \cref{E-transmission}, I found that this it is possible by developing two models that can infer incidence and transmission from prevalence surveys.
These did both rely on knowledge of the distribution of the duration of detectability, which can partially be inferred from the CIS.
One model was phenomenological in nature and the other mechanistic.
The phenomenological approach, based on the backcalculation framework, provides estimates of incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
The mechanistic approach, based on an SEIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
The choice of modelling approach should depend on the situation.
A phenomenological approach is more robust when there are high levels of uncertainty; for example, early in an outbreak or in the presence of new variants the infectious period duration may not be known.
On the other hand, a mechanistic approach provides more information.
Where possible, both methods should be implemented; discrepancies between results may highlight deficiencies in modelling assumptions.

@@

A final consideration is the estimation of the severity of a pathogen.
Severity of a pathogen is crucial for risk assessment and can change over time~\autocite[e.g.][]{kirwanSeverityTrends}.
Two approaches could be taken to using a prevalence surveys for this purpose.
First, a cohort approach where the outcomes of individuals who test positive are followed up.
The low number of positive cases in the CIS would make this challenging.
In addition, the data would need to be linked to relevant records.
Second, a transmission model, such as the one in \cref{E-SEIR}, could jointly model severe events and prevalence.
Such an approach has previously been used~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2} but without the sophistication of the observation model developed in \cref{E-SEIR:sec:observation}.

>>>>

In \cref{E-transmission}, I found that this it is possible by developing two models that can infer incidence and transmission from prevalence surveys.
These did both rely on knowledge of the distribution of the duration of detectability, which can partially be inferred from the CIS.
One model was phenomenological in nature and the other mechanistic.
The phenomenological approach, based on the backcalculation framework, provides estimates of incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
The mechanistic approach, based on an SEIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
The choice of modelling approach should depend on the situation.
A phenomenological approach is more robust when there are high levels of uncertainty; for example, early in an outbreak or in the presence of new variants the infectious period duration may not be known.
On the other hand, a mechanistic approach provides more information.
Where possible, both methods should be implemented; discrepancies between results can highlight deficiencies in modelling assumptions.

@@

A final consideration is the estimation of the severity of a pathogen.
Severity of a pathogen is crucial for risk assessment and can change over time~\autocite[e.g.][]{kirwanSeverityTrends}.
Two approaches could be taken to using a prevalence surveys for this purpose.
First, a cohort approach where the outcomes of individuals who test positive are followed up.
The low number of positive cases in the CIS would make this challenging.
In addition, the data would need to be linked to relevant records.
Second, a mechanistic model, such as the one in \cref{E-SEIR}, could jointly model severe events and prevalence.
Such an approach has previously been used~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2} but without the sophistication of the observation model developed in \cref{E-SEIR:sec:observation}.
